---
title: "EJ_Screen_Maeve"
author: "Maeve"
format: html
editor: visual
---

## Preamble

```{r}
#reading in libraries 
library(tidyverse) 
library(sf) 
library(here) 
library(stars) # for raster data
library(tmap) # for static and interactive maps 
library(viridis) 
#loading data (Census block group level) 
ejscreen <- sf::st_read(here::here("data","ejscreen","EJSCREEN_2023_BG_StatePct_with_AS_CNMI_GU_VI.gdb")) 
#Filtering only for AK
ejscreen_ak <- ejscreen %>% filter(ST_ABBREV == "AK")
```

## Making Map 1 (Percentile of Low Life Expectancy Households for Alaska)

```{r}
tm_shape(ejscreen_ak) + 
  tm_polygons( fill = "P_LIFEEXPPCT", 
               fill.scale = tm_scale_continuous(values = magma(8)), 
               fill.legend = tm_legend( title = "Percentile of Low Life Expectancy Households by Census tract", position = tm_pos_out("right","center") ) ) + 
        tm_graticules() + 
        tm_title_out("Percentile of Low Life Expectancy Households In Alaska By Census tract", 
        position = tm_pos_out("center","top"))
```

## Making Map 2 (Percentile of Low income households by Census Tract in Alaska)

```{r}
tm_shape(ejscreen_ak) + 
  tm_polygons( fill = "P_LESSHSPCT", 
               fill.scale = tm_scale_continuous(values = magma(8)), 
               fill.legend = tm_legend( title = "Census tract Percentile", position = tm_pos_out("right","center") ) ) + 
        tm_graticules() + 
        tm_title_out("Percentile of people with less than a High School Education in Alaska", 
        position = tm_pos_out("center","top"))
```

## What do the maps communicate?

I made two maps of Alaska, both are organized by Census tract, using EJScreen data from the EPA. The first describes the percentile of low life expectancy households for the state, with 0 being the highest concentration of low life expectancy households in the state, and 100 being the lowest. The second map describes the concentration of households in a tract with less than a highschool education, organized by percentile. These maps depict a relationship between low life expectancy and educational access, as many of the tracts which rank highly for life expectancy also rank highly for educational access, and vis versa. What is likely happening is that groups who have low educational access have reduced access to health care or other life-extending resources, which reduces their life expectancy for the household. \## Data Citation: EPA, 2024, "Environmental Justice Mapping and Screening Tool (EJScreen)", https://doi.org/10.7910/DVN/RLR5AX, Harvard Dataverse, V4, UNF:6:Ew64oHBMGoTrNkLoYBJcUw== \[fileUNF\]

------------------------------------------------------------------------
